# CS544 Module 6 Assignment
## Student: David J Lotito


#################
# Part1) Strings
#################

library(stringr)

# Use the stringr functions for the following:
# Initialize the vector of words from Lincoln’s speech with the following code:

file <- "http://people.bu.edu/kalathur/datasets/lincoln.txt"
words <- scan(file, what=character())

# a) Detect and show all the words that have a punctuation symbol.

words[str_detect(words, "[:punct:]")]

# b) Replace all the punctuations in the words dataset with an empty string. Remove any resulting empty words from this dataset. Convert all the remaining words to lower case.

remove_punct <- str_replace_all(words, "[:punct:]", "")

remove_empty_string <- remove_punct[remove_punct != ""]

# Make this the new_words dataset.
new_words <- str_to_lower(remove_empty_string)

# c) What are the top 5 frequent words in the new_words dataset?

names(sort(table(new_words), decreasing = T))[1:5]


# d) Show the frequencies of the word lengths in the new_words dataset. Plot the distribution of these frequencies.
``` {r, echo = F}

hist(str_length(new_words), xlim = c(1, 15))

```

# e) What are the words in the new_words dataset with the longest length?

max(str_length(new_words))

new_words[str_detect(new_words, "^[a-zA-Z]{16}$")]

# f) Show all the words in the new_words dataset that start with the letter p.

new_words[str_detect(new_words, "^[p]")]

# g) Show all the words in the new_words dataset that end with the letter r.

new_words[str_detect(new_words, "[r]$")]

# h) Show all the words in the new_words dataset that start with the letter p and end with the letter r.

new_words[str_detect(new_words, "^[p]+[a-zA-Z]+[a-zA-Z]+[r]$")]

# In c), you realize that the most spoken words are what are known as stopwords. In text mining, the stopwords are removed before analysis. Initialize the common English stopwords as follows:

stopfile <- "http://people.bu.edu/kalathur/datasets/stopwords.txt" 
stopwords <- scan(stopfile, what=character())

# Remove the stopwords from the new_words data. Use the %in% operator. Repeat c) and d) for this dataset without the stop words.

new_words_WO_stop_words <- new_words[!(new_words %in% stopwords)]

names(sort(table(new_words_WO_stop_words), decreasing = T))[1:5]

``` {r, echo = F}

hist(str_length(new_words_WO_stop_words), xlim = c(1, 15))

```

#######################
# Part2) Data Wrangling
#######################


















