---
title: "UFC Fights Analysis"
author: "David Lotito"
date: 9/28/2022
output:
  html_document:
    fig_width: 8
    code_folding: hide
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=8, fig.height=4,
                      echo=FALSE, warning=FALSE, message=FALSE)
```

# Dataset Overview

The following dataset is from Kaggle and contains the UFC-Fight historical data from 1993 to 2021.This dataset is a list of every UFC fight in the history of the organisation. Every row contains information about both fighters, fight details and the winner. The data was scraped using beautifulsoup in Python and processed via Pandas. This is the preprocessed file where each row contains information about each fighter collected up until this particular fight and the descriptive fight details. In this file, feature engineering, missing data treatment and one hot encoding are already done. The raw files have been merged to form this dataset.

# Data Prep

Each row is a compilation of both fighter stats. Fighters are represented by 'red' and 'blue' (for red and blue corner). So for instance, red fighter has the complied average stats of all the fights except the current one. The stats include damage done by the red fighter on the opponent and the damage done by the opponent on the fighter (represented by 'opp' in the columns) in all the fights this particular red fighter has had, except this one as it hasn't occured yet (in the data). Same information exists for blue fighter. The target variable is 'Winner' which is the only column that tells you what happened.

# Goal of Analysis

The goal of this analysis is to understand how the fights have changed over the course of time and to understand the fighters strategies

```{r echo=TRUE}
# Import CSV
ufc_data <- read.csv("/Users/davelotito/Code/Masters_Degree_Projects/Foundations of Analytics and Data Visualization/Term_Project/UFC_Preprocessed_Fight_Data.csv", header = T)
```


```{r echo=TRUE}
winner_title <- ufc_data[(c("Winner", "title_bout"))]

table(winner_title)

```


```{r, echo = FALSE}
barplot(table(winner_title), col = c("Blue", "Red"), xlab = "Title Bout vs. Non-Title Bout", ylab = "Number of Fights", main = "Wins by Color - Title Fight vs. Non-Title Fight")
```



















