---
title: "UFC Fights Analysis"
author: "David Lotito"
date: 9/28/2022
output:
  html_document:
    fig_width: 8
    code_folding: hide
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=8, fig.height=4,
                      echo=FALSE, warning=FALSE, message=FALSE)
```

```{r echo=TRUE}
# Import CSV
library(plotly)
library(tidyr)
library(prob)
library(sampling)

ufc_data <- read.csv("/Users/davelotito/Code/Masters_Degree_Projects/Foundations of Analytics and Data Visualization/Term_Project/UFC_Preprocessed_Fight_Data.csv", header = T)

winner_title <- ufc_data[(c("Winner", "title_bout"))]

```

# Dataset Overview

The following dataset is from Kaggle and contains the UFC-Fight historical data from 1993 to 2021. This dataset is a list of EVERY UFC fight in the history of the organisation. Every row contains information about both fighters, fight details and the winner. The data was scraped using beautifulsoup in Python and processed via Pandas. This is the preprocessed file where each row contains information about each fighter collected up until this particular fight and the descriptive fight details. In this file, feature engineering, missing data treatment and one hot encoding are already done. The raw files have been merged to form this dataset.

# Data Prep

Each row is a compilation of both fighter stats. Fighters are represented by 'red' and 'blue' (for red and blue corner). So for instance, red fighter has the complied average stats of all the fights except the current one. The stats include damage done by the red fighter on the opponent and the damage done by the opponent on the fighter (represented by 'opp' in the columns) in all the fights this particular red fighter has had, except this one as it hasn't occured yet (in the data). Same information exists for blue fighter. The target variable is 'Winner' which is the only column that tells you what happened.

# Goal of Analysis

The goal of this analysis is to understand how long fighters tend to last in the UFC, who typically wins more often(Red or Blue), and if you were to bet on fights based on color of the fighters corner(Red or Blue) would you be more likely to win on a specific color.

# Wins by Color

We want to understand if a specific color gives any indicator if a fighter is more likely to win a fight. In the below bar plot we are showing the frequency of wins for Blue vs. Red fighter color.

```{r, echo = FALSE}

barplot(table(winner_title[1]), col = c("Blue", "Red"), xlab = "Red Wins vs. Blue Wins", ylab = "Frequency", main = "Fights 1993-2021", ylim = c(0,5000))

```

### Wins by Color for Title Fights vs. Non-Title Fights

Since we can see that more red colored fighters are winning fights I wanted to see if the same applies when we add in another variable, in this case we are looking at fighting for a title. The results show that when fighting for a title, red fighters tend to win more often than blue fighters.


```{r, echo = FALSE}
barplot(table(winner_title), col = c("Blue", "Red"), xlab = "Title Bout vs. Non-Title Bout", ylab = "Frequency", main = "Fights 1993-2021 by Title Bout")
```
# Win Streak: Blue vs. Red

Now that we know that red fighters tend to win more often, both in title-fights and non-title fights I wanted to analyze what the winning streak looks like for both red and blue fighters. The analysis tells us that for both red fighters tend to have longer winning streaks vs. blue fighters. This makes sense as we saw from the other analysis we did on wins. The distributions are similar, but we can clearly see that red fighters tend to have longer winning streaks vs. blue fighters. Both plots are right skewed as evident from the plots provided.

``` {r, echo = FALSE}

win_streak_blue <- ufc_data[c("B_current_win_streak")]
win_streak_red <- ufc_data[c("R_current_win_streak")]

par(mfrow=c(1,2))   
 
barplot(table(win_streak_blue), col = c("Blue"), , xlab = "Blue # of Wins in Steak", ylab = "Frequency", main = "Duration of Win Streak: Blue Fighters", ylim = c(0, 3500))
barplot(table(win_streak_red), col = c("Red"), xlab = "Red # of Wins in Steak", ylab = "Frequency", main = "Duration of Win Streak: Red Fighters", ylim = c(0, 3000))

summary(win_streak_blue)
summary(win_streak_red)

```
# Distribution of Data
## Fighter Ages: Blue vs. Red

How long do fighters last in the UFC? Whats the average age of fighters? What do the outliers tell us? These are all questions that I wanted to analyze. As we can see from the analysis in the scatter plot that the majority of the fights are in their early 20s to late 30s indicating that as the fighters age they are less and less likely to continue fighting, likely due to the wear and tear on the body. We do have some outliers, of course, who fight well into their 40s and even 50s. The average age for the fighters overall is ~30. This makes sense as most professional athletes peak in their early to mid 30s.

```{r, echo = FALSE}
ages <- ufc_data[c("B_age", "R_age")]

plot(ages, main = "Age of Fighters - 1993 to 2021", col = c("blue", "red"))
legend(x=48,y=45,c("B_age","R_age"),cex=.8,col=c("blue","red"),pch=c(1,1))

summary(ages)
```

## Fighter Wins/Losses: Blue vs. Red

After analyzing the fighters number of wins, number of wins in a title fight, and the ages of fighters I wanted to take a deeper diver into the overall picture of the blue vs. red wins & losses to see the bigger picture. As we can see from the box plot below that the red fighters are clearly the better fighters overall with more wins over the blue fighters. Blue fighters have less losses compared to Red Fighters, but we need to take into account that there has been less fights overall for the blue fighters as we can see from the data provided.


```{r, echo = FALSE}
wins <- ufc_data[c("B_wins", "R_wins", "B_losses", "R_losses")]
boxplot(wins, horizontal = T, col=c("blue","red"), main = "Red/Blue: Wins vs. Losses", xlab = "Values", las = 1, cex.axis=0.8)
colSums(wins)
```
# Central Limit Theorem

The central limit theorem states that the distribution of sample means, taken from independent random sample sizes, follows a normal distribution even if the original population is not normally distributed. This is important because there are a lot of statistical procedures that require normality in the data set. As a result we can apply statistical techniques that assume normality even when the population is non normal. Using the age attribute in this data set the applicability of the central limit theorem can be shown. As displayed in the box plot and histogram above, the age distribution of all groups have a positive skew.

``` {r, echo = F}
set.seed(9386)

# Combine Ages into One Column
ages<- data.frame(ages = c(ufc_data[,"B_age"], ufc_data[,"R_age"]))

ages <- as.matrix(ages)


par(mfrow = c(2,2))

xbar <- c()

for (size in c(10, 20, 30, 40)) {
for (i in 1:nrow(ages)) {
xbar[i] <- mean(sample(x = ages, size = 10, replace = TRUE))
}
cat("Sample Size = ", size, " Mean = ", mean(xbar)," SD = ", sd(xbar), "\n")
hist(xbar, prob = TRUE, breaks = 15, main = paste("Sample Size =", size))

}


```

# Sampling

Sampling is a technique to select a representative portion of the population to perform different studies on. There are many different sampling techniques that can be leveraged in analytics including simple random sampling, systematic sampling, and stratified sampling. Simple random sampling is a basic sampling technique where individual subjects are selected from a larger group. In this case, every sample has the same chance of getting picked. Systematic sampling is a method where samples are selected via a fixed periodic interval. The interval is calculated by dividing the whole population sample by the desired sample size. The first sample is decided randomly within the first interval. Lastly, stratified sampling takes into the account that there is heterogeneity in a population. The population is subdivided into sub populations and the same percentage of individuals is selected from each sub population to make up the sample set. When looking at a normal distribution, the sample mean can be used as an estimate for the population mean. Given a certain confidence level, a confidence interval is defined. The confidence interval is range of values which contains the population mean with the given confidence level.

For this project the ages of ufc fighters population with be analyzed. Simple random sampling without replacement & systematic sampling will be utilized as sampling methods.

## Population of Ages
This simple histogram shows the distribution of the ages of all UFC fighters since inception of the sport in 1993.
``` {r, echo = F}

hist(ages, main = "Population of Ages", ylim = c(0, 2500))

```

## SRSWOR
Simple random sampling is when a specified sample is selected from the larger group or larger frame. Each person or object has an equivalent opportunity of getting selected. An example is if we have 4 firefighters, 6 cops, and 4 nurses and simple random sampling were to be used to pick a sample size of 5 any five the group of 14 could be picked.

``` {r, echo = F}

set.seed(9386)

srswor <- srswor(100, nrow(ages))

rows <- (1:nrow(ages))[srswor!=0]
rows <- rep(rows, srswor[srswor != 0])
sample.1 <- ages[rows, ]

df <- table(sample.1)

barplot(df, main = "SRSWOR", xlab = "Ages")

paste("SRSWOR Mean:", mean(df))
paste("SRSWOR SD:", sd(df))
```

## Systematic Sampling
Systematic sampling is when there are “rules” decided to pick the sample size. For example, if there were 10 houses on 4 streets. The “rule” could be that the sampling will start with the 2nd house and then every 3rd house will be chosen. There is a potential bias with this type of sampling

``` {r, echo = F}
set.seed(9386)

N = nrow(ages)
n = 100
k = ceiling(N / n)
r = sample(k, 1)
s = seq(r, by = k, length = n)

sample.2 <- ages[s, ]

df <- table(sample.2)

barplot(df, main = "Systematic Sampling", xlab = "Age")

paste("Systematic Sampling Mean:", mean(df))
paste("Systematic Sampling SD:", sd(df))

```


